Physcraper: a python package for continual update of evolutionary estimates using the Open Tree of Life

### 2020-07-27 docs/submission-mee-2020-07-27.tex
intro clean 4

\textbf{Word count}: 3431

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Phylogenies are a key part of research in many areas of biology. Tools that automatize
  some parts of the process of phylogenetic reconstruction, mainly molecular character matrix assembly,
  have been developed for the advantage of both specialists in the field of phylogenetics and nonspecialists.
  However, interpretation of results, comparison with previously available phylogenetic
  hypotheses, and choosing of one phylogeny for downstream analyses and discussion still impose difficulties
  to one that is not a specialist either on phylogenetic methods or on a particular group of study.
\item
  Physcraper is an open‐source, command-line Python program that automatizes the update of published
  phylogenies by enriching underlying gene alignments with public DNA sequence data, and linking taxonomic information across databases.
  This provides a framework for comparison of published phylogenies with their updated versions, by using the conflict Application Programming Interface (API) function of the Open Tree of Life project.
\item
  Physcraper can be used by the nonspecialist, as a tool to generate phylogenetic
  hypotheses based on already available expert phylogenetic knowledge.
  Phylogeneticists and group specialists will find it useful as a tool to facilitate molecular dataset gathering and comparison
  of alternative phylogenetic hypotheses (topologies).
\item
  We hope that the Physcraper workflow demonstrates the benefits of doing open science for phylogenetics, encouraging more researchers to strive for better sharing practices. Physcraper can be used with any OS and is released under an open source license. Detailed instructions for installation and
  use are available at \href{https://physcraper.readthedocs.io/en/tutorial/index.html}{https://physcraper.readthedocs}.
\end{enumerate}

@@

Phylogenetic estimates of evolutionary relationships capture the shared history of living organisms, and provide key context for all our biological observations.
Public biological databases constitute an amazing resource for evolutionary estimation, but a large portion of molecular data publicly available has never been incorporated into any phylogenetic estimate. Extending existing phylogenetic estimates with DNA sequence data from molecular databases in a reproducible and continuous manner is possible by automatizing connections between biological databases. Here, we propose Physcraper, a tool that takes advantage of existing phylogenetic research and data from public biological databases to update a starting tree and single locus alignments, leveraging on molecular data that taxon specialists have assessed and deemed appropriate for a specific phylogenetic scope.

One of the main challenges to automatically connect data from disparate biological databases in a phylogenetic context is the prevalence of taxonomic idiosincracies across databases. To standardize taxonomic names, a unified taxonomy is necessary. The main aim of the Open Tree of Life (OpenTree from now on) project
is to construct a comprehensive, dynamic and digitally-available tree of life by synthesizing published phylogenetic trees along with taxonomic data. Currently, OpenTree's ``synthetic'' tree comprises 2.3 million tips, of which around 90,000 are represented by phylogenetic estimates - the remaining 1.4 million taxa are placed in the tree based on their taxonomic assignment. To achieve this, OpenTree unifies/integrates taxonomy from various databases (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}) such as the USA National Center for Biodiversity Information (NCBI) molecular database GenBank (Benson \emph{et al.} \protect\hyperlink{ref-benson2000genbank}{2000}; Wheeler \emph{et al.} \protect\hyperlink{ref-wheeler2000database}{2000}), the Global biodiversity Information Facility(GBIF; Secretariat \protect\hyperlink{ref-secretariat2017gbif}{2017}), the Encyclopedia of Life (EOL; Wilson \protect\hyperlink{ref-wilson2003encyclopedia}{2003}), etc., providing a key resource that can be used to connect data from all these databases to phylogenetic data that has been standardized to OpenTree's unified taxonomy.

Another challenge to incorporate molecular data from public databases to update our phylogenetic knowledge is producing high-quality homology hypotheses. While the genomic era has allowed us to obtain large quantities of molecular data from a single species, different research questions call for different genomic sequencing approaches, from whole genomes, to transcriptomes, RadSeq, SNP, mirosats and UCE's.
Having a variety of approaches has resulted in largely non-overlapping genomic data sets across taxa. Moreover, even when the same sequencing approach is applied, such as RadSeq, phylogenetic distance can cause allelic dropout at deeper divergences (Eaton \emph{et al.} \protect\hyperlink{ref-eaton2016misconceptions}{2016}). In contrast, decades of single locus sequencing have generated homologous DNA data sets across large phylogenetic scales. Indeed, GenBank release number 159 (April 15, 2007) hosted 72 million DNA sequences that were gauged to have the potential to resolve phylogenetic relationships of most of its 241 000 distinct taxa (about 98.05\% of taxa in the NCBI taxonomy release 159; Sanderson \emph{et al.} \protect\hyperlink{ref-sanderson2008phylota}{2008}).

The TreeBASE database (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}) hosts about 8, 200 curated \emph{non-genomic?} alignments, providing information on evolutionary relationships of around 100, 000 distinct taxa \href{https://www.treebase.org/treebase-web/home.html\#:~:text=TreeBASE\%20is\%20produced\%20and\%20governed,mapped\%20to\%20104\%2C593\%20distinct\%20taxa.}{(see TreeBASE's website about)},
constituting an untapped source of valuable expert knowledge with the potential to update phylogenetic relationships in several different regions of the tree of life.

The OpenTree Phylesystem has incorporated all phylogenies from TreeBASE, while saving metadata linking trees to the corresponding alignment repository in TreeBASE. If there are multiple alignments and trees, TreeBASE does not always indicate how they are connected. Hence, this metadata only provides a loose means of linking the tree with the exact alignment that generated it and a human curator must make the link.

By linking molecular data, available from databases such as the GenBank (Benson \emph{et al.} \protect\hyperlink{ref-benson2000genbank}{2000}; Wheeler \emph{et al.} \protect\hyperlink{ref-wheeler2000database}{2000}), to alignments and phylogenies, available in
the TreeBASE repository (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}) and OpenTree's Phylesystem, we can place new biological data in an evolutionary context.

@@

@@

@@

@@

>>>>

\textbf{Word count}: 3652

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Phylogenies are a key part of research in many areas of biology. Tools that automatize
  some parts of the process of phylogenetic reconstruction, mainly molecular character matrix assembly,
  have been developed for the advantage of both specialists in the field of phylogenetics and nonspecialists.
  However, interpretation of results, comparison with previously available phylogenetic
  hypotheses, and choosing of one phylogeny for downstream analyses and discussion still impose difficulties
  to one that is not a specialist either on phylogenetic methods or on a particular group of study.
\item
  Physcraper is a command-line Python program that automatizes the update of published
  phylogenies by enriching underlying gene alignments with public DNA sequence data. It builds upon tools from the Open Tree of Life project to link taxonomic information across databases, providing a framework for straightforward comparison of published phylogenies with their updated versions.
\item
  Physcraper can be used by the nonspecialist, as a tool to generate phylogenetic
  hypotheses based on publicly available expert phylogenetic knowledge.
  Phylogeneticists and group specialists will find it useful as a tool to facilitate molecular dataset gathering and comparison
  of alternative phylogenetic hypotheses (topologies).
\item
  We hope that the Physcraper workflow demonstrates the benefits of doing open science for phylogenetics, encouraging more researchers to strive for better sharing practices. Physcraper can be used with any OS and is released under an open-source license. Detailed instructions for installation and
  use are available at \href{https://physcraper.readthedocs.io/en/tutorial/index.html}{https://physcraper.readthedocs}.
\end{enumerate}

@@

Phylogenetic estimates of evolutionary relationships capture the shared history of living organisms, and provide key context for all our biological observations.
Public biological databases constitute an amazing resource for evolutionary estimation, but a large portion of molecular data publicly available has never been incorporated into any phylogenetic estimate. Extending existing phylogenetic estimates with new DNA sequence data, geographical location, fossil data, etc., in a reproducible and continuous manner is possible by automatizing connections between biological databases. Here, we propose Physcraper, a tool that uses existing phylogenetic research and data from public biological databases to update a starting tree and single locus alignments, to build upon molecular data that taxon specialists have assessed and deemed appropriate for a specific phylogenetic scope.

The prevalence of taxonomic idiosincracies across databases represent a big challenge to automatically connect data from disparate biological databases in a phylogenetic context. To standardize taxonomic names, a unified taxonomy is needed. The main aim of the Open Tree of Life (OpenTree from now on) project
is to construct a comprehensive, dynamic and digitally-available tree of life by synthesizing published phylogenetic trees along with taxonomic data. Currently, OpenTree's ``synthetic'' tree comprises 2.3 million tips, of which around 90,000 are represented by phylogenetic estimates - the remaining 1.4 million taxa are placed in the tree based on their taxonomic assignment. To achieve this, OpenTree unifies taxonomy from various databases (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}) such as the USA National Center for Biodiversity Information (NCBI) molecular database GenBank (Benson \emph{et al.} \protect\hyperlink{ref-benson2000genbank}{2000}; Wheeler \emph{et al.} \protect\hyperlink{ref-wheeler2000database}{2000}), the Global Biodiversity Information Facility (GBIF; Secretariat \protect\hyperlink{ref-secretariat2017gbif}{2017}), and the Encyclopedia of Life (EOL; Wilson \protect\hyperlink{ref-wilson2003encyclopedia}{2003}), providing a key resource that can be used to connect data from virtually any biological database to phylogenetic data that has been standardized to OpenTree's unified taxonomy.

Another challenge to incorporating molecular data from public databases to update phylogenetic knowledge is assembling high-quality homology hypotheses. Species tree reconstructions from multi-gene data sets taking into account the multispecies coalescent model are seen as the gold standard for inferring species relationships (Song \emph{et al.} \protect\hyperlink{ref-song2012resolving}{2012}).
While genomics have allowed us to obtain large quantities of molecular data from one organism, different research questions call for different genomic sequencing approaches, from whole genomes, to transcriptomes, restriction-site associated DNA sequencing (RADSeq), single nucleotide polymorphisms (SNPs), microsatellites, and ultra-conserved elements (UCEs).
Having a variety of approaches has resulted in largely non-overlapping genomic data sets across taxa. Moreover, some types of genomic data are not suitable for phylogenetic inference at any scale. For example, phylogenetic distance can cause allelic dropout at deeper divergences with RADSeq data sets (Eaton \emph{et al.} \protect\hyperlink{ref-eaton2016misconceptions}{2016}).
In contrast, decades of single locus sequencing have generated homologous DNA data
that can be used for phylogenetic reconstruction at many scales. Indeed, more than a decade ago, GenBank release number 159 (April 15, 2007) already hosted 72 million non-genomic DNA sequences that were gauged to have the potential to resolve phylogenetic relationships of most of the 241, 000 distinct taxa in the NCBI taxonomy at the time (about 98.05\% of taxa; Sanderson \emph{et al.} \protect\hyperlink{ref-sanderson2008phylota}{2008}). Assembling a DNA alignment from such a massive database can be done ``by hand'', but it requires huge amounts of time and it is mostly a non-reproducible method. Computer pipelines that make DNA sequence search faster and more efficient, as well as more reproducible have been utilized to study evolutionary relationships among a variety of organisms (e.g., Smith \emph{et al.} \protect\hyperlink{ref-smith2009mega}{2009}; Izquierdo-Carrasco \emph{et al.} \protect\hyperlink{ref-izquierdo2014pumper}{2014}; Antonelli \emph{et al.} \protect\hyperlink{ref-antonelli2017toward}{2017}).
Still, there is a large tradition in phylogenetics of building molecular character matrices of few, thoughtfully curated markers, as manual curation of alignments often improves phylogenetic reconstructions (Fragoso-Martínez \emph{et al.} \protect\hyperlink{ref-fragoso2017pilot}{2017}).

A way to incorporate the best of two worlds (massive amounts of newly released molecular data and fine-grained curation from human experts) is to rely on published manually curated homology hypotheses as ``jump-start'' alignments (Morrison \protect\hyperlink{ref-morrison2006multiple}{2006}). The TreeBASE database (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}) hosts about 8, 200 publicly accesible alignments, providing information on evolutionary relationships of around 100, 000 distinct taxa \href{https://www.treebase.org/treebase-web/home.html\#:~:text=TreeBASE\%20is\%20produced\%20and\%20governed,mapped\%20to\%20104\%2C593\%20distinct\%20taxa.}{(see TreeBASE's website about)}, representing an untapped source of valuable expert knowledge with the potential to continuously enrich and update phylogenetic relationships in several different regions of the tree of life, by incorporating data from molecular databases that has not been included in any public phylogenetic estimate. The phylogenies associated to all these alignments have been integrated to the OpenTree's datastore, the Phylesystem (McTavish \emph{et al.} \protect\hyperlink{ref-mctavish2015phylesystem}{2015}), and metadata linking these trees to their corresponding alignment repository in TreeBASE is available.

We propose Physcraper as a Python encoded pipeline that relies on programmatic access protocols to link molecular data from GenBank, to alignments from TreeBASE and phylogenies from OpenTree's Phylesystem, to automatically update and enrich phylogenetic knowledge based on expertly-curated homology hypotheses. Physcraper also provides new types of access to various general OpenTree programmatic tools for custom comparison of phylogenetic hypotheses.
It focuses on database interoperability, by automatizing taxonomic name matching across biological databases, so users can perform any downstream analysis straightforwardly, as well as placing new data from virtually any biological database in an evolutionary context.

@@

\leavevmode\hypertarget{ref-antonelli2017toward}{}%
Antonelli, A., Hettling, H., Condamine, F.L., Vos, K., Nilsson, R.H., Sanderson, M.J., Sauquet, H., Scharn, R., Silvestro, D., Töpel, M. \& others. (2017). Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of taxa. \emph{Systematic Biology}, \textbf{66}, 152--166.

@@

\leavevmode\hypertarget{ref-izquierdo2014pumper}{}%
Izquierdo-Carrasco, F., Cazes, J., Smith, S.A. \& Stamatakis, A. (2014). PUmPER: Phylogenies updated perpetually. \emph{Bioinformatics}, \textbf{30}, 1476--1477.

@@

\leavevmode\hypertarget{ref-morrison2006multiple}{}%
Morrison, D.A. (2006). Multiple sequence alignment for phylogenetic purposes. \emph{Australian Systematic Botany}, \textbf{19}, 479--539.

@@

\leavevmode\hypertarget{ref-smith2009mega}{}%
Smith, S.A., Beaulieu, J.M. \& Donoghue, M.J. (2009). Mega-phylogeny approach for comparative biology: An alternative to supertree and supermatrix approaches. \emph{BMC evolutionary biology}, \textbf{9}, 37.

\leavevmode\hypertarget{ref-song2012resolving}{}%
Song, S., Liu, L., Edwards, S.V. \& Wu, S. (2012). Resolving conflict in eutherian mammal phylogeny using phylogenomics and the multispecies coalescent model. \emph{Proceedings of the National Academy of Sciences}, \textbf{109}, 14942--14947.
